Gut microbes and their genes are associated with cognition and neuroanatomy in children

### 2023-03-25 manuscript/main.tex
add supplementary table numbers

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Supplementary Figure 1)
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-abundance \emph{Prevotella copri} (Supplementary Figure 2).
Overall subject-to-subject variation in gut microbial genes
and brain volume profiles was similarly driven largely by subject age (Figure 1E-F).

@@

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.13, mean RMSE 12.70),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 17.29). For both age groups,
species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (Figure
3A-B, Supplementary Table \hl{XX}A-B); aside from except for \emph{Eubacterium ramulus}
all taxa that were significant using LMs belonged to a group
responsible for 80\% of the total relative importance in RF models
(Table 3, Figure 3A).
However, the rank-ordering of LM significance and RF importance
were had substantive differences. For example, in children under 6 months,
only \emph{B. wexlerae} was significantly associated with cognitive function in LMs, but
was ranked 20th in importance for RF models.

@@

If there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
employed a RF modeling approach to associate gut taxonomic
profiles with individual brain regions identified in MRI scans.
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 4, Supplementary
Table \hl{XX}), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.421, Right = 0.434; relative age importances, Left = 7.5\%, Right =
8.3\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.200, Right = 0.273; relative age importances, Left = 3.7\%, Right =
6.3\%). In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.288; relative age importances = 1.2\%).

@@

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a
consistent number of species between 14 and 22 (median = 20) was
responsible for one third of the fitness-weighted cumulative importance,
regardless of model performance (\hl{Suplementary Table XX}, Figure 4 A and B).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included seven species - \emph{Anaerostipes hadrus}, \emph{Bacteroides
vulgatus}, \emph{Fusicatenibacter saccharivorans}, \emph{Ruminococcus
torques}, \emph{Eubacterium rectale}, \emph{Coprococcus comes} and
\emph{Blautia wexlerae} - that combined, account for approximately one
third of the cumulative relative importances, computed after subsetting
on the taxa of interest (Figure 4C).

@@

The use of multiple age-appropriate cognitive assessments that could be
normalized to a common scale enabled us to analyze microbial
associations across multiple developmental periods, but carries several
drawbacks. In particular, the test-retest reliability, as well as
systematic differences between test administrators may introduce
substantial noise into these observations, particularly in the youngest
children. In addition, our study period overlapped with the beginning of
the COVID-19 pandemic, and we and others have observed some reduction in
measured scores for children that were assessed after the implementation
of lockdowns. In our subject set for this study, these effects are more
pronounced in some age groups due to the period when active recruitment for
the study was occurring.
\cite{blackwellYouthWellbeingCOVID192022,deoniImpactCOVID19Pandemic2021}
(Supplementary Figure 3).

>>>>

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Figure S1)
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-abundance \emph{Prevotella copri} (Figure S2).
Overall subject-to-subject variation in gut microbial genes
and brain volume profiles was similarly driven largely by subject age (Figure 1E-F).

@@

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.13, mean RMSE 12.70),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 17.29). For both age groups,
species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (Figure
3A-B, Table S1A-B); aside from except for \emph{Eubacterium ramulus}
all taxa that were significant using LMs belonged to a group
responsible for 80\% of the total relative importance in RF models
(Table 3, Figure 3A).
However, the rank-ordering of LM significance and RF importance
were had substantive differences. For example, in children under 6 months,
only \emph{B. wexlerae} was significantly associated with cognitive function in LMs, but
was ranked 20th in importance for RF models.

@@

If there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
employed a RF modeling approach to associate gut taxonomic
profiles with individual brain regions identified in MRI scans.
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 4, Table S2), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.421, Right = 0.434; relative age importances, Left = 7.5\%, Right =
8.3\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.200, Right = 0.273; relative age importances, Left = 3.7\%, Right =
6.3\%). In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.288; relative age importances = 1.2\%).

@@

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a
consistent number of species between 14 and 22 (median = 20) was
responsible for one third of the fitness-weighted cumulative importance,
regardless of model performance (Figure 4A-B, Table S2).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included seven species - \emph{Anaerostipes hadrus}, \emph{Bacteroides
vulgatus}, \emph{Fusicatenibacter saccharivorans}, \emph{Ruminococcus
torques}, \emph{Eubacterium rectale}, \emph{Coprococcus comes} and
\emph{Blautia wexlerae} - that combined, account for approximately one
third of the cumulative relative importances, computed after subsetting
on the taxa of interest (Figure 4C).

@@

The use of multiple age-appropriate cognitive assessments that could be
normalized to a common scale enabled us to analyze microbial
associations across multiple developmental periods, but carries several
drawbacks. In particular, the test-retest reliability, as well as
systematic differences between test administrators may introduce
substantial noise into these observations, particularly in the youngest
children. In addition, our study period overlapped with the beginning of
the COVID-19 pandemic, and we and others have observed some reduction in
measured scores for children that were assessed after the implementation
of lockdowns. In our subject set for this study, these effects are more
pronounced in some age groups due to the period when active recruitment for
the study was occurring.
\cite{blackwellYouthWellbeingCOVID192022,deoniImpactCOVID19Pandemic2021}
(Figure S3).
